---
title: "Operations on sequences"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Operations on sequences}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bioseq)
```

## Biological operations

### Transcribe and translate
- seq_transcribe
- seq_rev_transcribe
- seq_translate
- seq_rev_translate

### Reverse and complement
- seq_reverse
- seq_complement


## String operations

### Crop
- seq_crop_pattern
- seq_crop_position

### Detect
- seq_detect_pattern

### Remove
- seq_remove_pattern
- seq_remove_position
- seq_remove_gap
- seq_remove_ambiguous

### Replace
- seq_replace_pattern
- seq_replace_position

### Split
- seq_split_pattern
- seq_split_position
- seq_split_stop
- seq_split_codon
- seq_split_kmer

### Extract
seq_extract_pattern
seq_extract_position

### Count
- seq_count_pattern
- seq_count_codon
- seq_count_stop

### Assemble
- seq_combine
- seq_duplicate
- seq_insert


## Stats operations
- seq_nseq
- seq_nchar
- seq_stat_prop
- seq_stat_gc

## Summarize operations

- seq_consensus(x, methods = c("chr_majority", "chr_ambiguity", "seq_centrality", "seq_majority"), weights)
- seq_cluster(x, threshold = 0.97, method)

## Simulations
- seq_generate(what = "DNA", n_seq, n_base, code = 1)
- seq_randomize(x)
- seq_mutate(x, freq)

## Apply and columns
seq_transpose
Apply any functions to rows/column. They are treated as separate character. To work on column strings see seq_transpose.
seq_col_apply(x, .f)
seq_row_apply(x, .f)

## Misc
- seq_disambiguate_IUPAC
- seq_spellout
